---
title: "riskCommunicator Manuscript Vignette"
author: "Jessica Grembi, Elizabeth Rogawski McQuade"
date: "`r Sys.Date()`"
output: 
    rmarkdown::html_vignette:
      number_sections: no
      toc: yes
vignette: >
  %\VignetteIndexEntry{Short Vignette: Included in Manuscript}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{tidyverse}
---

```{r, echo = FALSE}
options(width = 60)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
  # tidy.opts = list(width.cutoff = 60,
  #                  args.newline = TRUE),
  # tidy = T
)
```

# Introduction to `riskCommunicator`

The `riskCommunicator` package facilitates the estimation of common epidemiological effect measures that are relevant to public health, but that are often not trivial to obtain from common regression models, like logistic regression. In particular, `riskCommunicator` estimates risk and rate differences, in addition to risk and rate ratios. The package estimates these effects using g-computation with the appropriate parametric model depending on the outcome (logistic regression for binary outcomes, Poisson regression for rate or count outcomes, negative binomial regression for overdispersed rate or count outcomes, and linear regression for continuous outcomes). Therefore, the package can handle binary, rate, count, and continuous outcomes and allows for dichotomous, categorical (>2 categories), or continuous exposure variables. Additional features include estimation of effects stratified by subgroup and adjustment of standard errors for clustering. Confidence intervals are constructed by bootstrap at the individual or cluster level, as appropriate. 

This package operationalizes g-computation, which has not been widely adopted due to computational complexity, in an easy-to-use implementation tool to increase the reporting of more interpretable epidemiological results. To make the package accessible to a broad range of health researchers, our goal was to design a function that was as straightforward as the standard logistic regression functions in R (e.g. `glm`) and that would require little to no expertise in causal inference methods or advanced coding.


# Getting started

## Installation  

The `riskCommunicator` R package is available from CRAN so can be installed using the following command:
```{r installation, eval = F}
install.packages("riskCommunicator")
```

Load packages:
```{r setup}
library(riskCommunicator)
library(tidyverse)
library(printr)
library(formatR)
library(sandwich)
library(stringr)
library(ggpubr)
```

## Description of main package function

The `gComp` function is the main function in the `riskCommunicator` package and allows you to estimate a variety of effects depending on your outcome and exposure of interest. The function is coded as follows:

```{r, printr.help.sections = c('usage','arguments')}
?gComp
```

# Framingham Heart Study

We'll demonstrate how to use the package with data from the Framingham Heart Study. The following information is from the official Framingham study documentation (https://biolincc.nhlbi.nih.gov/teaching/):

"The Framingham Heart Study is a long term prospective study of the etiology of cardiovascular disease among a population of free living subjects in the community of Framingham, Massachusetts. The Framingham Heart Study was a landmark study in epidemiology in that it was the first prospective study of cardiovascular disease and identified the concept of risk factors and their joint effects. The study began in 1948 and 5,209 subjects were initially enrolled in the study. Participants have been examined biennially since the inception of the study and all subjects are continuously followed through regular surveillance for cardiovascular outcomes. Clinic examination data has included cardiovascular disease risk factors and markers of disease such as blood pressure, blood chemistry, lung function, smoking history, health behaviors, ECG tracings, Echocardiography, and medication use. Through regular surveillance of area hospitals, participant contact, and death certificates, the Framingham Heart Study reviews and adjudicates events for the occurrence of Angina Pectoris, Myocardial Infarction, Heart Failure, and Cerebrovascular disease.

```{r}
data(cvdd)
```

**cvdd** is a subset of the data collected as part of the Framingham study from 4,240 participants who conducted a baseline exam and were free of prevalent coronary heart disease when they entered the study.
Participant clinic data was collected during three examination periods, approximately 6 years apart, from roughly 1956 to 1968. Each participant was followed for a total of 24 years for the outcome of the following events: Angina Pectoris, Myocardial Infarction, Atherothrombotic Infarction or Cerebral Hemorrhage (Stroke) or death. 

NOTE: This is a "teaching" dataset. Specific methods were employed to ensure an anonymous dataset that protects patient confidentiality; therefore, this dataset is inappropriate for publication purposes." The use of these data for the purposes of this package were approved on 11Mar2019 (request #7161) by NIH/NHLBI.


## Binary outcome example

Research question: what is the effect of having diabetes at the beginning of the study on the 24-year risk of cardiovascular disease or death due to any cause? 

Here, we will estimate the risk difference, risk ratio, odds ratio, and number needed to treat, adjusting for patient's age, sex, body mass index (BMI), smoking status (current smoker or not), and prevalence of hypertension (if they are hypertensive or not at baseline). Logistic regression is used as the underlying parametric model for g-computation.

```{r binary_outcome, paged.print = FALSE, cache=T}
## Specify the regression formula
cvdd.formula <- cvd_dth ~ DIABETES + AGE + SEX + BMI + CURSMOKE + PREVHYP

## For reproducibility, we should always set the seed since the g-computation uses  
## random resampling of the data to calculate confidence intervals and random  
## sampling of the distribution when predicting outcomes.
set.seed(1298)

## Call the gComp function
binary.res <- gComp(data = cvdd, 
                    formula = cvdd.formula, 
                    outcome.type = "binary", 
                    R = 1000)

binary.res
```

The result obtained from the `gComp` function is an object of class **gComp** which is a list containing the summary results, `results.df`, `n`, `R`, `boot.result`, `contrast`, `family`, `formula`, `predicted.outcome`, and `glm.result` (see `?gComp` or `help(gComp)` for a more detailed explanation of each item in the list).  

```{r gComp_class_explaination}
class(binary.res)

## The names of the different items in the list: 
names(binary.res)

## Sample size of the original data:
binary.res$n 

## Contrast being compared in the analysis:
binary.res$contrast
```

### Checking model fit

The 95% CIs obtained from the riskCommunicator package represent population-standardized marginal effects obtained with g-computation. To ensure that the parameter estimates from each bootstrap iteration are normally distributed, we can also look at the histogram and Q-Q plots of bootstrapped estimates by calling:
```{r binary_outcome_plot, fig.width = 12, fig.height = 9, out.width = "100%"}
plot(binary.res)
```
The histograms show the different effect estimates obtained by each bootstrap resampling of the data and should be normally distributed if the model is correctly specified.  Q-Q plots help to verify that the bootstrap values are normally distributed by comparing the actual distribution of bootstrap values against a theoretical normal distribution of values centered at mean = 0. If the estimates are normally distributed, the plotted estimates (black circles) should overlay the diagonal red dashed line.

### In the manuscript, we compare the results of gComp to standard regression models. Here, we show how to do that so we can re-create Table 2.

First we obtain the odds ratio using logistic regression.
```{r std-regression-odds-ratio, echo = F}
std.reg.or = glm(formula = cvd_dth ~ DIABETES + AGE + SEX + BMI + CURSMOKE + PREVHYP, 
                 data = cvdd, 
                 family=binomial(link='logit'))

std.reg.or

df.std.reg.or = data.frame(exp(cbind(OR = coef(std.reg.or), confint.default(std.reg.or, level = 0.95)))) %>%
  # and we pull out only the estimate we are concerned with for the predictor DIABETES
  filter(rownames(.) == "DIABETES1") %>%
  rename(LL = `X2.5..`, UL = `X97.5..`)

```


Next, we calculate the risk ratio using a Poisson approximation of log-binomial regression with robust variance (sandwich standard errors).
```{r std-regression-risk-ratio, echo = F}
std.reg.rr = glm(formula = cvd_dth ~ DIABETES + AGE + SEX + BMI + CURSMOKE + PREVHYP, 
                 ## to use Poisson regression, we need to change DIABETES from a factor to a numeric (0,1) variable
                 data = cvdd %>%
                   mutate(cvd_dth = ifelse(cvd_dth == "0", 0, ifelse(cvd_dth == "1", 1, NA))), 
                 family="poisson")
std.reg.rr

## calculate robust Standard errors
cov.std.reg.rr = sandwich::vcovHC(std.reg.rr)
std.err <- sqrt(diag(cov.std.reg.rr))

df.std.reg.rr = data.frame(Estimate = exp(coef(std.reg.rr)[2]), 
                      Robust_SE = exp(std.err[2]),
                      LL = exp(coef(std.reg.rr)[2] - 1.96 * std.err[2]), 
                      UL = exp(coef(std.reg.rr)[2] + 1.96 * std.err[2]))

```


We can try to calculate the risk difference using a log-linear regression, but the model won't converge.
```{r std-regression-risk-difference, eval = F}
std.reg.rd = glm(formula = cvd_dth ~ DIABETES + AGE + SEX + BMI + CURSMOKE + PREVHYP, 
                 data = cvdd %>%
                   ## To use linear regression, we need to change DIABETES from a 
                   ##   factor to a numeric (0,1) variable
                   mutate(cvd_dth = ifelse(cvd_dth == "0", 0, 
                                           ifelse(cvd_dth == "1", 1, NA))), 
                 family=gaussian(link = 'log'))
```

We can re-create Table 2 from the manuscript now!
```{r table 2}
# combine standard regression results
std.regression.res = df.std.reg.or %>%
  mutate(Parameter = "Odds Ratio",
         Std_regression = paste0(round(OR, 2), 
                                 " (", 
                                 round(LL, 2), 
                                 ", ", 
                                 round(UL, 2), 
                                 ")")) %>%
  bind_rows(df.std.reg.rr %>%
              mutate(Parameter = "Risk Ratio",
                     Std_regression = paste0(round(Estimate, 2), 
                                             " (", 
                                             round(LL, 2), 
                                             ", ", 
                                             round(UL, 2), 
                                             ")"))) %>%
  select(Parameter, Std_regression) %>%
  rename(`Standard regression` = Std_regression)
rownames(std.regression.res) = NULL


(table2 = binary.res$results.df %>%
  mutate(riskCommunicator = paste0(format(round(Estimate, 2), 2), 
                                   " (", 
                                   format(round(`2.5% CL`, 2), 2), 
                                   ", ", 
                                   format(round(`97.5% CL`, 2), 2), 
                                   ")"),
         riskCommunicator = ifelse(Parameter == "Number needed to treat/harm", 
                                   round(Estimate, 2), riskCommunicator)) %>%
  select(Parameter, riskCommunicator) %>%
  left_join(std.regression.res, by = "Parameter"))
```


## Rate outcome example

Research question: what is the effect of having diabetes at the beginning of the study on the rate of cardiovascular disease or death due to any cause? 

Here, we will estimate the rate difference and rate ratio, adjusting for patient's age, sex, body mass index (BMI), smoking status (current smoker or not), and prevalence of hypertension (if they are hypertensive or not at baseline). We have included **timeout** as the `offset` and a `rate.multiplier` of `365.25*100` so that the estimates are returned with units of 100 person-years. Poisson regression is used as the underlying parametric model for g-computation. (Note: for overdispersed count/rate outcomes, the negative binomial distribution can be specified by setting `outcome.type` to **"count_nb"** or **"rate_nb"**.)


```{r rate_outcome, cache = T}
## Modify the dataset to change the variable cvd_dth from a factor 
##  to a numeric variable since the outcome for Poisson
##  regression must be numeric.
cvdd.t <- cvdd %>%
  dplyr::mutate(cvd_dth = as.numeric(as.character(cvd_dth)),
                timeout = as.numeric(timeout))

set.seed(6534)

rate.res <- gComp(data = cvdd.t, 
                  Y = "cvd_dth", 
                  X = "DIABETES", 
                  Z = c("AGE", "SEX", "BMI", "CURSMOKE", "PREVHYP"), 
                  outcome.type = "rate", 
                  rate.multiplier = 365.25*100, 
                  offset = "timeout", 
                  R = 1000)

rate.res
```
  

## Rate outcome with subgroups example

Research question: what is the effect of having diabetes at the beginning of the study on the rate of cardiovascular disease or death due to any cause, stratified by sex? 

Here, we will estimate the same effects above, but in subgroups defined by sex.


```{r rate_outcome_subgroup, cache = T}

rate.res.subgroup <- gComp(data = cvdd.t, 
                           Y = "cvd_dth", 
                           X = "DIABETES", 
                           Z = c("AGE", "SEX", "BMI", "CURSMOKE", "PREVHYP"), 
                           subgroup = "SEX", 
                           outcome.type = "rate", 
                           rate.multiplier = 365.25*100, 
                           offset = "timeout", 
                           R = 1000)

rate.res.subgroup

```
  
### The `results.df` component of the `gComp` function output is formatted as a data.frame.  This makes it very easy to immediately plot the results using **ggplot2** or your favorite R plotting functionality.  Here's an example for plotting the results of the different subgroups (sexes) for the rate example above.

```{r plot-rate-outcome-results, warning = F}
# Saving the output and modifying the labels of the SEX variable to specify 
#   male/female instead of 0/1
# Also adding a new variable to show the line indicating no difference found 
#   (0 for rate difference, 1 for rate ratio)

df = rate.res.subgroup$results.df %>%
  mutate(Subgroup = ifelse(Subgroup == "SEX0", "Male", "Female"),
         hline = ifelse(Parameter == "Incidence Rate Ratio", 1, 0)) 

ggplot(df, aes(x = Subgroup, y = Estimate)) + 
  geom_point(size = 2) + 
  geom_errorbar(aes(ymin = `2.5% CL`, ymax = `97.5% CL`), 
                width = 0.2) + 
  facet_wrap(~Parameter) +
  theme_bw() + 
  labs(x = "", color = "") + 
  geom_hline(aes(yintercept = hline), 
             color = "red", 
             linetype = "dashed", 
             alpha = 0.3) 
```

### Let's make a figure comparing the results we got above using riskCommunicator with those we get using standard regression models for this rate example.

First, we need to get the covariate-conditional estimates using standard Poisson regression.
```{r std-regression-models-rate}
# Standard Poisson regression (spr) for the rate question, across both sexes
spr.rate = glm(
  formula = cvd_dth ~ DIABETES + AGE + SEX + BMI + CURSMOKE + PREVHYP + 
    offset(log(timeout+0.001)),
  data = cvdd.t, 
  family = "poisson"
  )

# get the parameter estimate and CI from the model object
df.spr.rate = as.data.frame(
  exp(cbind(Estimate = coef(spr.rate), confint.default(spr.rate, level = 0.95)))
  ) %>%
  rename(`2.5% CL` = `2.5 %`, `97.5% CL` = `97.5 %`) %>%
  filter(rownames(.) == "DIABETES1") %>%
  mutate(Subgroup = "All")


# Standard Poisson regression (spr) for the rate question, by subgroup (SEX)
spr.rate.subgroup = glm(
  formula = cvd_dth ~ DIABETES + AGE + SEX + BMI + CURSMOKE + PREVHYP + 
    DIABETES*SEX + offset(log(timeout+0.001)), 
  data = cvdd.t, 
  family = "poisson"
  )

# Get the variance-covariance matrix so we can calculate standard errors
se.subgroup = vcov(spr.rate.subgroup)

# Get estimates and CIs
df.spr.rate.subgroup = data.frame(
  Subgroup = c("Male", "Female"),
  raw.est = c(coef(spr.rate.subgroup)[2], 
              coef(spr.rate.subgroup)[2] + coef(spr.rate.subgroup)[8]),
  SE = c(sqrt(se.subgroup[2,2]), 
         sqrt(se.subgroup[2,2] + se.subgroup[8,8] + 2*se.subgroup[2,8]))) %>%
  mutate(Estimate = exp(raw.est), 
         `2.5% CL` = exp(raw.est - 1.96 * SE),
         `97.5% CL` = exp(raw.est + 1.96 * SE))

# Combine the results from the subgroup and full model into a single data.frame
combined.std.reg = df.spr.rate %>%
  bind_rows(df.spr.rate.subgroup %>%
            select(Subgroup, Estimate:`97.5% CL`)) %>%
  mutate(Parameter = "Incidence Rate Ratio",
         model = "Standard Poisson Regression")
```

Now, we can plot the same figure shown in the manuscript.
```{r plot-rate-outcome-results-v-std-regression, warning = F, fig.width = 8, fig.height = 4}
# Combine the riskCommunicator results with the standard Poisson regression results
df.combined = rate.res$results.df %>%
  mutate(Subgroup = "All") %>%
  bind_rows(df) %>%
  mutate(model = "riskCommunicator") %>%
  select(-Outcome, -Comparison) %>%
  bind_rows(combined.std.reg) %>%
  mutate(hline = ifelse(Parameter == "Incidence Rate Ratio", 1, 0))


rate.diff = ggplot(df.combined %>% 
                      filter(Parameter == "Incidence Rate Difference"), 
                   aes(x = Subgroup, y = Estimate, color = model)) + 
  geom_point(size = 2, position = position_dodge(width = .5)) + 
  geom_errorbar(aes(ymin = `2.5% CL`, ymax = `97.5% CL`), 
                width = 0.2, 
                position = position_dodge(width = .5)) + 
  scale_color_manual(values = c("#481567FF", "#3CBB75FF")) + 
  theme_bw() + 
  labs(x = "",  
       y = str_wrap('Incidence rate difference of 
                    cardiovascular disease or death 
                    (cases/100 person-years)', width = 32),
       color = "") + 
  geom_hline(aes(yintercept = hline), 
             color = "red", 
             linetype = "dashed", 
             alpha = 0.3) + 
  theme(legend.position = "none") 

rate.ratio = ggplot(df.combined %>% 
                       filter(Parameter == "Incidence Rate Ratio"), 
                   aes(x = Subgroup, y = Estimate, color = model)) + 
  geom_point(size = 2, position = position_dodge(width = .5)) + 
  geom_errorbar(aes(ymin = `2.5% CL`, ymax = `97.5% CL`), 
                width = 0.2, 
                position = position_dodge(width = .5)) + 
  scale_y_continuous(trans = "log2") +
  scale_color_manual(values = c("#481567FF", "#3CBB75FF")) + 
  theme_bw() + 
  labs(x = "",  
       y = str_wrap('Incidence rate ratio of 
                    cardiovascular disease or death 
                    (shown on natural log scale)', width = 32),
       color = "") + 
  geom_hline(aes(yintercept = hline), 
             color = "red", 
             linetype = "dashed", 
             alpha = 0.3) + 
  theme(legend.position = "bottom") 

ggarrange(rate.ratio, rate.diff, ncol = 2, common.legend = T, legend = "bottom",
          labels = c("A", "B"), widths = c(1, 1))
```

```{r save-Fig1, include = F, eval = F}
ggsave(filename = "~/Dropbox/Framingham teaching data set/Manuscript drafts/Fig1.tiff", device = "tiff", width = 10, height = 6)
```


```{r}
sessionInfo()
```
